Wildcard: Spreadsheet-Driven Customization of Web Applications

### 2020-01-10 paper.tex
done for the day

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Please go to https://dl.acm.org/ccs/ccs.cfm and generate your Classification
% System [view CCS TeX Code] stanz and copy _all of it_ to this place.
%% From HERE
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10002944.10011122.10003459</concept_id>
<concept_desc>General and reference~Computing standards, RFCs and guidelines</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10010405.10010476.10010477</concept_id>
<concept_desc>Applied computing~Publishing</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[300]{General and reference~Computing standards, RFCs and guidelines}
\ccsdesc[500]{Applied computing~Publishing}

% To HERE
%%%%%%%%%%%%%%%%%%%%%%%

@@

Web applications often don't match the particular needs of their users.
Sometimes there is a browser extension available to patch an issue, and
if the user is a programmer they might be able to fix it themselves. But
for most people, the only recourse is to complain to the developers, or
more likely, to simply give up. Back in 1977, in \emph{Personal Dynamic
Media} \autocite{kay1977}, Alan Kay envisioned personal computing as a
medium that let a user ``mold and channel its power to his own needs,''
but today, software feels more like concrete than clay.

In this paper, we present \emph{spreadsheet-driven customization}, a
technique that enables end users to customize software. The idea is to
augment an application's UI with a spreadsheet that is synchronized with
the application's data. When the user manipulates the spreadsheet, the
underlying data is modified and the changes are propagated to the UI,
and vice versa. We have implemented this technique in a prototype
browser extension called Wildcard. Section~\ref{sec:demos} presents
demos of using Wildcard to augment real websites in useful ways, and
Section~\ref{sec:implementation} describes the implementation of the
extension.

@@

Spreadsheet-driven customization provides an easy entry point for end
users, since small tweaks can be performed with a single click. At the
same time, it also supports a variety of richer customizations, like
adding private annotations to a webpage or joining in related data from
a web API. In Section~\ref{sec:design-principles}, we elaborate on this
principle of ``low floor, high ceiling,'' as well as other design
principles guiding our work.

@@

@@

The Wildcard extension is currently an early research prototype. We plan
to continue testing the system with our own use cases to understand
better how well the spreadsheet abstraction maps to real websites and
customization needs. Eventually we plan to release the tool publicly, in
order to study how end users choose to customize applications, discover
usability challenges, and to test the feasibility of programmers
building and maintaining site adapters.

@@

To get a sense of the user experience of using Wildcard, let's see an
example of someone using it to help with booking a trip.{ These demos
are best viewed as videos in the online version of this paper
(\url{https://www.geoffreylitt.com/wildcard}).}

@@

In 2012, the travel site Airbnb removed the ability to sort
accommodation listings by price. Users could still filter by price
range, but could no longer view the cheapest listings first. Many users
complained on online message boards that the change seemed hostile to
users. ``It's so frustrating!..What is the logic behind not having this
function?'' said one user on the
\href{https://community.withairbnb.com/t5/Hosting/Sorting-listing-by-price/td-p/559404}{Airbnb
support forum}. Alas, the feature remains missing to this day.

@@

Then, the user clicks on the price column header to sort the
spreadsheet, and the Airbnb UI, by in ascending order by price{
(Figure~\ref{fig:airbnb-demo}B)}. They can similarly use the column
headers to filter to listings with a user rating above 4.5, which the
original Airbnb UI also doesn't allow.

@@

Most websites that show lists of data also offer actions on rows in the
table, like adding an item to a shopping cart. Wildcard has the ability
to make these ``row actions'' available in the data table through the
site adapter.

In the Airbnb UI, saving multiple listings to a Favorites list requires
tediously clicking through them one by one. Using Wildcard, the user can
select multiple rows and favorite all of them with a single click{
(Figure~\ref{fig:airbnb-demo}D)}. Similarly, the user can also open the
detailed pages for many listings at once.

Now the user wants to jot down some notes on the pros and cons of each
listing. To do this, they type some notes into an additional column in
each listing row, and the notes appear inside the listings in the
original UI{ (Figure~\ref{fig:airbnb-demo}A)}. The annotations are saved
in the browser for future sessions.

Wildcard also includes a formula language, which enables more
sophisticated customizations that fetch external data and perform
computations.

When traveling without a car, it's useful to evaluate potential places
to stay based on how walkable the surroundings are. Using Wildcard
formulas, the user can integrate Airbnb with Walkscore, an API that
rates the walkability of any location on a 1-100 scale. When they type
in a \texttt{WALKSCORE} formula with the latitude and longitude of the
listing, it returns the walk score as the cell's value. Because the
cell's contents are injected into the page, the score also shows up in
the page body{ (Figure~\ref{fig:airbnb-demo}C)}.

@@

In previous examples the table cells were read-only; users cannot change
the name or price of an Airbnb listing. In this next case, the cells are
writable, so that changes in the table are reflected in the form inputs.
This becomes especially useful when combined with GUI widgets that can
edit the value of a table cell.

Filling in dates for a flight search typically requires opening up a
separate calendar app to find the right dates and then manually copying
them into the form. In Wildcard, the user can avoid this cumbersome
workflow by using a datepicker widget that includes the user's personal
calendar information{ (Figure~\ref{fig:expedia-demo}B)}. The user can
directly click on the right date, and it gets inserted into the
spreadsheet and the original form.

Here we've presented just a few use cases for spreadsheet-driven
customization, but we think the interactive data table is flexible
enough to support a wide range of other useful modifications, all while
remaining familiar and easy to use.

@@

In order to promote extensibility, Wildcard is implemented as a small
core program along with several types of plugins: site adapters,
formulas, and cell renderers/editors. The core contains functionality
for displaying the data table and handling user interactions, and the
table implementation is built using the
\href{https://handsontable.com/}{Handsontable} Javascript library.

@@

Wildcard provides an interface for concisely expressing how the DOM
should be mapped to a data table. For example, here is a code snippet
for extracting the name of an Airbnb listing:

@@

Sometimes important data is not shown in the UI, making it impossible to
scrape from the DOM. To address this, we have also prototyped mechanisms
for site adapters to observe AJAX requests made by the browser and
extract data directly from JSON responses. This mechanism was used to
implement the Airbnb Walkscore example, since latitude and longitude
aren't shown in the Airbnb UI, but they are available in AJAX responses.
In the future we also plan to add the ability for site adapters to
scrape data across multiple pages in paginated lists of results (as
explored in prior work \autocite{huynh2006}).

The site adapter also needs to support the reverse direction: sending
updates from the table to the original page. Most DOM manipulation is
not performed directly by the site adapter. Instead, Wildcard
automatically mutates the DOM to reflect the table state, using the same
declarative spec shown above. The only exception is row actions (like
favoriting an Airbnb listing), which are implemented as imperative
Javascript functions that can can mutate the DOM, simulate clicks on
buttons, etc.

@@

Computing does not need to be organized this way. For example, UNIX
offers a compelling alternative design: many small single-purpose
utilities, all of which manipulate a universal format of text streams.
The universal format creates a high degree of leverage from tools: users
can get a lot of utility from deeply mastering a text editor and some
text manipulation utilities, because these tools can be applied to
nearly any task. As just one example, a user's preferred text editor can
even serve as an interactive input mechanism in shell programs, e.g.~for
editing git commit messages.

Spreadsheet-driven customization aims to bring some of this UNIX ethos
to the world of isolated Web applications, by creating a consistent data
structure to represent the data inside many applications. In UNIX, the
universal format is a text stream; in Wildcard, it is a relational
table. The table is a simple abstraction which is also generic enough to
describe the data used in many different applications. Because Wildcard
maps the data from all applications to the table format, users can
invest in mastering the Wildcard table editor, the formula language, and
cell editor UIs, and reuse those same tools to customize many different
applications.

This idea relates to Beaudouin-Lafon and Mackay's notion of
\emph{polymorphic interface instruments} \autocite{beaudouin-lafon2000}:
UI elements that can be used in different contexts, like a color picker
that can be used in many different drawing applications. diSessa has
also noted the connection between literacy and the genericness of a
medium. Textual literacy rests on the fact that writing can be adapted
to many different genres and uses \autocite{disessa2000}; if people
needed to relearn reading and writing from scratch when switching from
essays to emails, the medium would lose most of its potency. We think
providing generic tools is especially important for software
customization, because the most common barrier to customizing software
is not having enough time \autocite{mackay1991}---it's more likely that
people will customize software regularly if they can reuse the same tool
across many applications.

This design principle leads to several challenges. First, any universal
abstraction has its constraints, and can't necessarily naturally express
the data in every application. We plan to explore the limits of the
table abstraction further, by trying to build adapters for more sites
with varied data formats. We expect that many types of data can fit
fairly naturally into tables: lists of search results, news articles,
and messages can all be seen as relations. On sites that use document
structures (e.g.~Google Docs) or graph structures (e.g.~social friend
graphs), it may prove more challenging to map internal data to this
abstraction.

Another challenge is ensuring a clear mapping in the user's mind between
the spreadsheet and the original page. Wildcard provides live visual
cues as the user navigates the data table (similar to the highlighting
provided by DOM inspectors in browser developer tools). In our own
usage, we have found that this live highlighting makes it very clear how
the two representations map to each other.

@@

One of the most interesting properties of spreadsheets is that users who
are only know a tiny sliver of their functionality (e.g., storing tables
of numbers or computing simple sums) can still use them in genuinely
valuable ways. The fact that useful tasks can be performed early on
supports the user's natural motivation to continue using the tool, and
to eventually learn its more powerful features if needed
\autocite{nardi1991}. In contrast, many traditional programming systems
require an enormous upfront investment of time and practice before
someone is able to write a program that helps them achieve a useful task
in their life.

@@

Real-world spreadsheet usage in offices is highly collaborative: most
users just perform simple changes, while a few coworkers help with
writing more complex formulas or even programming macros
\autocite{nardi1990}. Inspired by this collaborative approach, we aim to
make spreadsheet-driven customization a collaborative activity that
combines the different abilities of many users.

The main way we do this is by separating website customization into two
separate stages: structured data extraction, and using the resulting
spreadsheet. The data extraction stage is currently only available to
programmers who can code site adapters in Javascript, whereas the
customization stage is available to any non-programmer end user. This
architecture frees end users from needing to think about data
extraction, and enables a community of end users to reuse the efforts of
programmers building site adapters.

The group of users building adapters does not necessarily need to be
limited only to programmers. In the future, we plan to explore enabling
end users to also create site adapters, drawing on related work in this
area \autocite{chasins2018,huynh2006}. But even in that case, we still
envision a separation between highly motivated, tech-savvy end users
building adapters, and more casual end users just using the spreadsheet
view.

Another stakeholder to consider is the first party developers of the
original software. Spreadsheet-driven customization does not depend on
cooperation from first-party website developers, but if they were to
expose structured data in their web application clients, it would make
site adapters unnecessary, producing more robust results with much less
work.

We think there are compelling reasons for first parties to consider
doing this. Providing Wildcard support would allow users to build
extensions to fulfill their own feature requests. It also would not
necessarily require much effort: adding Wildcard support could be made
fairly straightforward for a first-party that already has direct access
to the structured data in the page. There is also precedent for first
parties implementing an official client extension API in response to
user demand: for several years, Google maintained an official extension
API in Gmail for Greasemonkey scripts to use.\footnote{Incidentally,
  since then, third parties have continued to maintain stable Gmail
  extension APIs used by many browser extensions
  \autocite{streak,talwar2019}, illustrating the potential of
  collaboratively maintaining third party adapters.}

@@

While similar in broad goals, Wildcard employs a different solution
strategy. These projects generally require building software from
scratch in a new environment, whereas Wildcard aims to maximize the
malleability of software built with existing tools. This approach has
the pragmatic benefit of enabling customization for much more software,
although it also imposes more rigid constraints.

With substantial future work, we think Wildcard could become more
similar to these other projects, growing from a platform for tweaking
existing software into a platform for building new software from
scratch. This would likely end up resembling existing tools for building
spreadsheet-driven applications (discussed more below), but with an
additional focus on customizability by end users of the software.

@@

Wildcard's goals are closely shared with other systems that provide
interfaces in the browser for end users to augment and customize
websites while using them.

@@

Sifter \autocite{huynh2006} enables users to sort and filter lists of
data on web pages, which resembles Wildcard's sort and filter
functionality. The underlying mechanism is also similar, since Sifter
extracts structured data from the page to enable its user-facing
functionality. The systems also have major differences. Wildcard aims to
extend this approach to support much broader functionality, and shows
the structured data table directly to the user. In contrast, Sifter only
supports sorting and filtering, and does not reveal the underlying data
table. The extraction mechanism is also different: Sifter uses a
combination of automated heuristics and interactive user feedback to
enable end users to extract data, whereas Wildcard hides data extraction
from end users.

Thresher \autocite{hogue2005} enables users to create wrappers which map
unstructured website content to Semantic Web content. Like Wildcard and
Sifter, Thresher augments the experience of original page based on
identifying structure; once semantic content has been identified, it
creates context menus in the original website which allow users to take
actions based on that content. Wildcard and Thresher focus on
complementary parts of the problem. Thresher aims to enable end users to
create content wrappers, but the actions available on the structured
data are created by programmers; conversely, Wildcard delegates wrapper
creation to programmers but gives end users more flexibility to use the
structured data in an open-ended way.

@@

``Sloppy programming'' \autocite{little2010} tools like Chickenfoot
\autocite{bolin2005} and Coscripter \autocite{leshed2008} enable users
to create scripts that perform actions like filling in text boxes and
clicking buttons, without directly interacting with the DOM. Users
express the desired page elements in natural, informal terms
(e.g.~writing ``the username box'' to represent the textbook closest to
the label ``username''), and then using heuristics to determine which
elements most likely match the user's intent. This approach allows for
expressing a wide variety of commands with minimal training, but it also
has downsides \autocite{little2010}:

\begin{itemize}
\tightlist
\item
  Reliability: It can be difficult to know whether a command will
  consistently work over time. In addition to changes to the website,
  changes to the heuristics can also cause problems.
\item
  Discoverability: it can be difficult for users to discover the space
  of possible commands.
\end{itemize}

Wildcard offers a sharp contrast to sloppy programming, instead choosing
to expose a high degree of structure through the familiar spreadsheet
table. Wildcard offers more consistency; for example, clicking a sort
header will always work correctly as long as the site adapter is
maintained. Wildcard also offers clearer affordances for what types of
actions are possible, or, crucially, what actions are \emph{not}
possible, which is useful to know. On the other hand, Wildcard's
explicit site adapter approach will result in coverage of many fewer
websites, and also a narrower range of possible actions than ``sloppy''
tools.

@@

Many others have made the powerful realization that a spreadsheet can
serve as an backing data store and computation layer for an interactive
web application, enabling end users to create such applications more
easily. Research projects like Object Spreadsheets
\autocite{mccutchen2016}, Quilt \autocite{benson2014}, Gneiss
\autocite{chang2014}, and Marmite \autocite{wong2007}, as well as
commercial tools like Airtable Blocks \autocite{zotero-79} and Glide
\autocite{zotero-81} allow users to view data in a spreadsheet table,
compute over the data using formulas, and then connect the table to a
GUI. Because many users are already familiar with using spreadsheets,
this way of creating applications tends to be much easier than
traditional software methods; for example, in a user study of Quilt,
many users were able to create applications in under 10 minutes, even if
they expected it would take them many hours.

@@

Web scraping tools differ in how much structure they attempt to map onto
the data. Some tools like Rousillon \autocite{chasins2018} extract data
in a minimally structured relational format; other tools like Piggy Bank
\autocite{huynh2005} more ambitiously map the data to a rich semantic
schema. In Wildcard, we chose to avoid schemas, in order to minimize the
work associated with creating a site adapter.

In the future, we might be able to integrate web scraping tools to help
create more reliable site adapters for Wildcard with less work, and to
open up adapter creation to end users. Sifter was built on top of the
Piggy Bank scraping library, suggesting precisely this type of
architecture where web scraping tools are used to support interactive
page modification.

@@

The most important question is: where are the limits of this
computational model? What types of useful customizations can it support
or not support? While initial demos suggest a variety of use cases, we
plan to develop more site adapters and demos to explore this question
further. We will start with privately testing the system with our own
needs, and then eventually deploy the tool publicly, once the API is
stable enough and can support a critical mass of sites and use cases. We
also plan to run usability studies to evaluate and improve the design of
the tool.

@@

A second potential limitation is that there is currently no mechanism
for end users to express imperative workflows with sequences of actions,
which related work has shown to be useful
\autocite{bolin2005,leshed2008}. It's not clear how such workflows could
fit into Wildcard, since spreadsheets have a fundamentally different
computation model, and site adapters created by programmers cannot
easily account for every possible action a user would want to take in a
page.

Another open question is how efficiently site adapters can be created
for real websites, which often include complex markup. We plan to
explore this question by creating adapters for sites with different data
structures and in different domains, and perhaps running user tests with
programmers. Possible future improvements we've considered include
developing automated heuristics to assist with the adapter creation
process, and developing new abstractions that make it easier for
programmers to efficiently create new robust adapters.

@@

\acks
Thank you to Glen Chiacchieri, Steve Krouse, Daniel Windham, and Maggie
Yellen for providing helpful feedback on this work.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Please go to https://dl.acm.org/ccs/ccs.cfm and generate your Classification
% System [view CCS TeX Code] stanz and copy _all of it_ to this place.
%% From HERE
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011066.10011069</concept_id>
<concept_desc>Software and its engineering~Integrated and visual development environments</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Integrated and visual development environments}
% To HERE
%%%%%%%%%%%%%%%%%%%%%%%

@@

Web applications often don't match the particular needs of their users.
Sometimes there is a browser extension available to patch an issue, and
if the user is a programmer they might be able to fix it themselves. But
for most people, the only recourse is to complain to the developers, or
more likely, to simply give up. Back in 1977, in \emph{Personal Dynamic
Media} \autocite{kay1977}, Alan Kay envisioned personal computing as a
medium that let a user ``mold and channel its power to his own needs,''
but today, software behaves more like concrete than clay.

In this paper, we present \emph{spreadsheet-driven customization}, a
technique that enables end users to customize software without doing any
traditional programming. The idea is to augment an application's UI with
a spreadsheet that is synchronized with the application's data. When the
user manipulates the spreadsheet, the underlying data is modified and
the changes are propagated to the UI, and vice versa. We have
implemented this technique in a prototype browser extension called
Wildcard. Section~\ref{sec:demos} presents demos of using Wildcard to
augment real websites in useful ways, and
Section~\ref{sec:implementation} describes the implementation of the
extension.

@@

@@

Spreadsheet-driven customization provides an easy entry point for end
users, since small tweaks like sorting data can be performed with a
single click. At the same time, it also supports a variety of richer
customizations, like adding private annotations to a webpage or joining
in related data from a web API. In Section~\ref{sec:design-principles},
we elaborate on this breadth of use cases, as well as other design
principles guiding our work.

@@

The Wildcard extension is currently an early research prototype. We plan
to continue testing the system with our own use cases to explore how the
spreadsheet abstraction maps to real websites and customization needs.
Eventually we plan to release the tool publicly, in order to study real
applications, discover usability challenges, and to test the feasibility
of programmers building and maintaining site adapters.

@@

To demonstrate the user experience of Wildcard, here is an example of
using it to help with booking a trip.{ These demos are best viewed as
videos in the online version of this paper
(\url{https://www.geoffreylitt.com/wildcard}).}

@@

In 2012, the travel site Airbnb removed the ability to sort
accommodation searches by price. Users could still filter by price
range, but could no longer view the cheapest listings first. Many users
complained that the change seemed hostile to users. ``It's so
frustrating!..What is the logic behind not having this function?'' said
one user on the
\href{https://community.withairbnb.com/t5/Hosting/Sorting-listing-by-price/td-p/559404}{Airbnb
support forum}. Alas, the feature remains missing to this day.

@@

Then, the user clicks on the price column header to sort the spreadsheet
and the Airbnb UI by price{ (Figure~\ref{fig:airbnb-demo}B)}. They also
filter to listings with a user rating above 4.5 (another feature missing
in the original Airbnb UI).

@@

Most websites that show lists of data also offer actions on rows in the
table, like adding an item to a shopping cart. Wildcard has the ability
to make these ``row actions'' available in the data table through the
site adapter. In the Airbnb UI, saving multiple listings to a Favorites
list requires tediously clicking through them one by one. Using Wildcard
row actions, the user can select multiple rows and favorite all of them
with a single click{ (Figure~\ref{fig:airbnb-demo}D)}. Similarly, the
user can open the detailed pages for many listings at once.

Next, the user wants to jot down some notes about each listing. To do
this, they type some notes into an additional column in each listing
row, and the notes appear inside the listings in the original UI{
(Figure~\ref{fig:airbnb-demo}A)}. The annotations are saved in the
browser for future sessions.

Wildcard also includes a formula language that enables more
sophisticated customizations. When traveling without a car, it's useful
to evaluate potential places to stay based on how walkable the
surroundings are. Using Wildcard formulas, the user can integrate Airbnb
with Walkscore, an API that rates the walkability of any location on a
1-100 scale. When they type in a \texttt{walkscore} formula with the
latitude and longitude of the listing, it sets the walk score as the
value of the cell. Because the cell's contents are injected into the
page, the score also shows up in the UI{
(Figure~\ref{fig:airbnb-demo}C)}.

@@

In previous examples the table cells were read-only, since users cannot
change the name or price of an Airbnb listing. In this case, the cells
are writable, which means that changes in the table are reflected in the
form inputs. This becomes especially useful when combined with GUI
widgets that can edit the value of a table cell.

Filling in dates for a flight search typically requires opening up a
separate calendar app to find the right dates and then manually copying
them into the form. In Wildcard, the user can avoid this cumbersome
workflow by using a datepicker widget that includes the user's personal
calendar events{ (Figure~\ref{fig:expedia-demo}B)}. The user can
directly click on the right date, and it gets inserted into the
spreadsheet and the original form.

Here we've presented just a few use cases for spreadsheet-driven
customization. We think the spreadsheet model is flexible enough to
support a wide range of other useful modifications, while remaining
familiar and easy to use.

@@

In order to promote extensibility, Wildcard is implemented as a small
core program along with several types of plugins: site adapters,
formulas, and cell editors. The core contains functionality for
displaying the data table and handling user interactions, and the table
implementation is built using the
\href{https://handsontable.com/}{Handsontable} Javascript library.

@@

Wildcard provides an interface for concisely expressing how the DOM
should be mapped to the spreadsheet table. For example, here is a code
snippet for extracting the name of an Airbnb listing:

@@

Sometimes sophisticated scraping techniques are necessary to extract the
necessary data from the application. For example, when we have
prototyped mechanisms for site adapters to observe AJAX requests made by
the browser and extract data directly from JSON responses. This
mechanism was used to implement the Airbnb Walkscore example, since
latitude and longitude aren't shown in the Airbnb UI, but they are
available in AJAX responses. This technique seems promising because AJAX
responses tend to already contain data in a structured form, and web
applications increasingly load data using AJAX. Another technique we
might consider adding is the ability for site adapters to scrape data
across multiple pages for paginated lists of results (as explored in
\autocite{huynh2006}).

The site adapter also needs to support the reverse direction: sending
updates from the table to the original page. Most DOM manipulation is
not performed directly by the site adapter. Instead, Wildcard
automatically mutates the DOM to reflect the spreadsheet state, using
the same declarative spec shown above. The only exception is row actions
(like favoriting an Airbnb listing), which are implemented as imperative
Javascript functions that can can mutate the DOM, simulate clicks on
buttons, etc.

@@

Computing does not need to be organized this way. For example, UNIX
offers a compelling alternative design: many small single-purpose
utilities, all of which manipulate a universal format of text streams.
The universal format creates a high degree of leverage from tools: users
can get a lot of utility from deeply mastering a text editor and text
manipulation utilities like \texttt{cut} and \texttt{sed}, because these
tools can be applied to a huge variety of tasks. A user's preferred text
editor can even serve as an interactive input mechanism in shell
programs, e.g.~for editing git commit messages.

Spreadsheet-driven customization aims to bring some of the UNIX ethos to
the world of isolated Web applications, by creating a consistent data
structure to represent the data inside many applications. In UNIX, the
universal format is a text stream; in Wildcard, it is a relational
table. Because Wildcard maps the data from all applications to the table
format, users can invest in mastering the Wildcard table editor, the
formula language, and cell editor UIs, and reuse those same tools to
customize many different applications.

This idea relates to Beaudouin-Lafon and Mackay's notion of
\emph{polymorphic interface instruments} \autocite{beaudouin-lafon2000},
which are UI elements that can be used in different contexts (for
example, a color picker that can be used in many different drawing
applications). It also relates to ideas of literacy in a medium. diSessa
notes that textual literacy rests on the fact that writing can be
adapted to many different genres and uses \autocite{disessa2000}; if
people needed to relearn reading and writing from scratch when switching
from essays to emails, the medium would lose most of its potency. Having
generic tools is especially important for software customization,
because the most common barrier to customizing software is not having
enough time \autocite{mackay1991}. It's more likely that people will
customize software regularly if they can reuse the same tools across
many applications.

This design principle leads to several challenges. First, any universal
abstraction has its constraints, and can't necessarily naturally express
the data in every application. We plan to explore the limits of the
table abstraction further, by trying to build adapters for more sites
with varied data formats. We expect that many types of data can fit
easily into tables: lists of search results, news articles, and messages
can all naturally be seen as relations. For sites that use document
structures (e.g.~Google Docs) or graph structures (e.g.~social
networks), it may prove more challenging to map internal data to the
table abstraction.

@@

One of the most interesting properties of spreadsheets is that users who
are only know a tiny sliver of their functionality (e.g., storing tables
of numbers or computing simple sums) can still use them in genuinely
valuable ways. The fact that useful tasks can be performed early on
supports the user's natural motivation to continue using the tool, and
to eventually learn its more powerful features if needed
\autocite{nardi1991}. In contrast, many traditional programming systems
require an enormous upfront time investment before someone is able to
write a program that helps them achieve a useful task in their life.

@@

Real-world spreadsheet usage is highly collaborative. Many users just
perform simple changes, while their coworkers help with writing more
complex formulas or even programming macros \autocite{nardi1990}.
Inspired by this, we aim to make spreadsheet-driven customization a
collaborative activity that combines the different abilities of
different users.

The main way we do this is by separating website customization into two
separate stages: structured data extraction, performed by programmers
who code site adapters in Javascript, and then customization using the
resulting spreadsheet, which is available to all end users. This
architecture frees end users from needing to think about data
extraction, and enables a community of end users to reuse the efforts of
programmers building site adapters.

The group of users building adapters does not necessarily need to be
limited only to programmers. In the future, we plan to explore enabling
end users to also create site adapters, drawing on related work in this
area \autocite{chasins2018,huynh2006}. But even then, we still envision
a separation between highly motivated, tech-savvy end users building
adapters, and more casual end users just using the spreadsheet view.

Another group to consider is the first party developers of the original
software. Spreadsheet-driven customization does not depend on
cooperation from first-party website developers, but if they were to
expose structured data in their web application clients, it would
eliminate the need for third party site adapters. We think there are
compelling reasons for first parties to consider doing this. Providing
Wildcard support would allow users to build extensions to fulfill their
own feature requests. It also would not necessarily require much effort,
since adding Wildcard support could be fairly straightforward for a
first-party that already has direct access to the structured data in the
page. There is also precedent for first parties implementing an official
client extension API: for several years, Google maintained an official
extension API in Gmail for Greasemonkey scripts to use.\footnote{Incidentally,
  since then, third parties have continued to maintain stable Gmail
  extension APIs used by many browser extensions
  \autocite{streak,talwar2019}, illustrating the potential of
  collaboratively maintaining third party adapters.}

@@

While similar in broad goals, Wildcard employs a different solution
strategy. These projects generally require building software from
scratch in a new environment, whereas Wildcard aims to maximize the
malleability of software built with existing tools.

With substantial future work, we think Wildcard could become more
similar to these other projects, growing from a platform for tweaking
existing software into a platform for building new software from
scratch. This would likely end up resembling existing tools for building
spreadsheet-driven applications (discussed more below), but with an
additional focus on customization by end users of the software.

@@

Wildcard's goals are closely shared with other systems that provide
interfaces in the browser for end users to customize websites.

@@

Sifter \autocite{huynh2006} enables users to sort and filter lists of
data on web pages, which resembles Wildcard's sorting functionality. The
underlying mechanism is similar, since Sifter extracts structured data
from the page to enable its user-facing functionality. However, the
systems also have significant differences:

\begin{itemize}
\tightlist
\item
  Wildcard aims support a broad set of customizations, but Sifter only
  supports sorting and filtering.
\item
  Wildcard shows the structured data table directly to the user. Sifter
  only shows sorting controls, and does not reveal the underlying data
  table.
\item
  Wildcard hides data extraction from end users. Sifter uses a
  combination of automated heuristics and interactive user feedback to
  involve end users in extracting data.
\end{itemize}

Thresher \autocite{hogue2005} enables users to create wrappers which map
unstructured website content to Semantic Web content. Like Wildcard and
Sifter, Thresher augments the page; once semantic content has been
identified, it creates context menus in the original website which allow
users to take actions based on that content. Wildcard and Thresher focus
on complementary parts of the customization process. Thresher aims to
enable end users to create content wrappers, but the actions available
on the structured data are determined in advance. Conversely, Wildcard
delegates wrapper creation to programmers, but gives end users more
flexibility to use the structured data in an open-ended way.

@@

``Sloppy programming'' \autocite{little2010} tools like Chickenfoot
\autocite{bolin2005} and Coscripter \autocite{leshed2008} enable users
to create web automation scripts without directly interacting with the
DOM. The scripts can perform actions like filling in text boxes and
clicking on buttons. Users express the desired page elements in natural,
informal terms (e.g.~writing ``the username box'' to represent the
textbox closest to the label ``username''), and then the system uses a
set of heuristics to determine which elements most likely match the
user's intent. This approach allows for expressing a wide variety of
commands with minimal training, but it also has downsides
\autocite{little2010}:

\begin{itemize}
\tightlist
\item
  \emph{Reliability}: It can be difficult to know whether a command will
  consistently work over time. In addition to changes to the website,
  changes to the heuristics can also cause problems.
\item
  \emph{Discoverability}: it can be difficult for users to discover the
  space of possible commands.
\end{itemize}

Wildcard offers a sharp contrast to sloppy programming, instead choosing
to expose a high degree of structure through the familiar spreadsheet
table. Wildcard offers more consistency; for example, clicking a sort
header will always work correctly as long as the site adapter is
maintained. Wildcard also offers clearer affordances for what types of
actions are possible, or, crucially, what actions are \emph{not}
possible, which is useful to know. On the other hand, Wildcard's
explicit site adapter approach means that fewer websites can be
customized, and also that users can't perform customizations if the
relevant data is not in the spreadsheet.

@@

Many others have had the powerful realization that a spreadsheet can
serve as an backing data store and computation layer for an interactive
web application, enabling end users to create such applications more
easily. Research projects like Object Spreadsheets
\autocite{mccutchen2016}, Quilt \autocite{benson2014}, Gneiss
\autocite{chang2014}, and Marmite \autocite{wong2007}, as well as
commercial tools like Airtable \autocite{zotero-79} and Glide
\autocite{zotero-81} allow users to view data in a spreadsheet table,
compute over the data using formulas, and then connect the table to a
GUI. Because many users are already familiar with using spreadsheets,
this way of creating applications tends to be much easier than
traditional software methods; for example, in a user study of Quilt,
many users were able to create applications in under 10 minutes, even if
they expected it would take them many hours.

@@

Web scraping tools differ in how much structure they attempt to map onto
the data. Some tools like Rousillon \autocite{chasins2018} extract data
in a minimally structured relational format; other tools like Piggy Bank
\autocite{huynh2005} more ambitiously map the data to a rich semantic
schema. In Wildcard, we chose to avoid semantic schemas, in order to
minimize the work associated with creating a site adapter.

In the future, we might try to integrate existing web scraping tools, to
help create more reliable site adapters for Wildcard with less work, and
to open up adapter creation to end users.

@@

The most important question is: what are the limits of this
computational model? What types of useful customizations can it support
or not support? While initial demos suggest a variety of use cases, we
plan to develop more site adapters and demos to explore this question
further. We will start with privately testing the system with our own
needs, and then eventually deploy the tool publicly, once the API is
stable enough and can support a critical mass of sites and use cases. We
also plan to run usability studies to evaluate and improve the design of
the tool.

@@

A second limitation is that there is currently no way for end users to
express imperative workflows with sequences of actions. Related work has
shown these types of workflows to be useful
\autocite{bolin2005,leshed2008} in practice. It's not clear how such
workflows could fit into Wildcard, since spreadsheets have a
fundamentally different computation model, and site adapters created by
programmers cannot easily account for every possible action a user would
want to take in a page.

Another open question is how easily site adapters can be created and
maintained for real websites, which often include complex markup and
change frequently. We plan to explore this question by creating adapters
for sites with different data structures and in different domains, and
perhaps running user tests with programmers. Possible future
improvements we've considered include developing automated heuristics to
assist with the adapter creation process, and developing new
abstractions that make it easier for programmers to efficiently create
new robust adapters.

@@

\acks
Thanks to Glen Chiacchieri, David Karger, Steve Krouse, Rob Miller, Arvind Satyanarayan, Daniel Windham, and Maggie Yellen for providing valuable feedback on this work.
